The Sloan Digital Sky Survey Spectroscopic Reduction Pipeline

### 2006-03-28 doc/paper/skysubtraction.tex
concurrent update

We utilize an optimal extraction procedure to transform a 2-dimensional image of
the fiber spectrograms to a set of 320 1-dimensional spectra.  We use the same procedure
on all processed frames, and will describe the details in processing the flat-field frames, the
arc lamp frames, and finally the science frames.  
The extraction is performed row-by-row, so each frame requires an outside loop over 2048 individual rows.
The counts in each row of 2048 pixels are modeled with a linear combination of 320 profiles plus a low-order
polynomial background.  We decided to perform profile fitting to the SDSS spectral images 
for the following reasons:

@@

The major obstacle to actual optimal extraction is the determination of the exact shape and position 
of the underlying profile.  We investigated a range of symmetric, analytic profiles with a 
minimum number of free parameters.  In the end, we found the best results using a simple gaussian 
or a slightly steeper profile characterized by $e^{-|x|^3/\sigma^3}$.  The profiles
are defined to be normalized to unity, so the amplitude of the profile fit gives a direct estimator
of the counts at a single spectral position.

Extraction is first carried out on all flat-field exposures (typically 10 second observations of
4 quartz lamps projected on petals covering the entrance aperture of the telescope) which were observed
with a common plug plate configuration.  The flat-field extraction is used to determine the shape of
the profile as a function of wavelength and position on the CCD.  The resulting spectral extractions are also used to 
determine the empirical fiber-to-fiber variations in throughput (as a function of wavelength) 
during a single setup.

The first very critical step is to robustly determine the fiber positions in CCD pixel coordinates.
This is performed by identifying peaks in a cross-sectional cut of across the center row (actually a median
cross-section of the central 9 rows).  The correct identification of the 320
fibers per frame can be uniquely determined by demanding a wider spacing between the bundles (about 10 pixels)
relative to the nominal inter-fiber spacing (6.1 pixels on average).  Once the starting peaks and 
identification have been determined, we trace the fiber centers by performing flux-weighted centroids on subsequent rows
using the previous row's center as the intial guess.  This has proven to work spectacularly well on SDSS
flat-fields, and we have yet to uncover a failure mode in over 10000 flat-field frames.
The flux-weighted centroids do have scatter from shot noise, so we fit a 7th order polynomial to each set of
fiber centroids, and the best-fit coefficients are stored to represent the fiber positions on each
flat-field frame.

Three types of exposures, corresponding to flat-field, arc lamp and science frames,
 are extracted with the same basic procedure
which we describe below.  The frames associated with a single observing sequence 
are processed in a certain order.  The first types to be extracted are the 
flat-fields frames ("FLATS").

The object extraction lists 5 steps in header:

>>>>

We utilize an optimal extraction procedure to transform a 2-dimensional image of
the fiber spectrograms to a set of 320 1-dimensional spectra (per spectrograph).
We use the same procedure on all processed frames, 
and will describe the details in processing the flat-field frames, the
arc lamp frames, and finally the science frames.  
The extraction is performed row-by-row, so each frame requires an 
outside loop over 2048 individual rows.
The counts in each row of 2048 columns are modeled with a linear 
combination of 320 profiles plus a low-order
polynomial background.  In addition, deviations to the trace centers and
the profile widths are calculated as linear basis modes (reflecting the first
and second spatial derivatives, respectively).  The compiled routine
we implement to fit the profiles in a single row (extract_row.c) can
be invoked with any combination of these parameters.

We decided to perform profile fitting to the 
SDSS spectral images for the following reasons:

@@

The major obstacle to actual optimal extraction is the determination of 
the exact shape and position of the underlying profile.  
We investigated a range of symmetric, analytic profiles with a 
minimum number of free parameters.  In the end, we found the best 
results using a simple Gaussian or a slightly steeper profile 
characterized by $e^{-|x|^3/\sigma^3}$.  The profiles are defined to be 
normalized to unity, so the amplitude of the profile fit is a 
direct estimator of the counts at a single spectral position.

Extraction is first carried out on all flat-field exposures 
(typically 10 second observations of 4 quartz lamps projected on petals covering the entrance aperture of the telescope) which were observed
with a common plug plate configuration.  The flat-field extraction is 
used to determine the shape of
the profile as a function of wavelength and position on the CCD.  
The resulting spectral extractions are also used to 
determine the empirical fiber-to-fiber variations in throughput 
(as a function of wavelength) during a single setup.   
A flat-field frame is labeled rejected by the pipeline if one of the
following conditions is met:  1) More the 2\% of the CCD pixels are labeled
bad by sdssproc, 2) More than 100 rows of the image show saturation, 
3) The 80th percentile count level is less than 1000 counts, 
4) Fewer than all 8 flat-field telescope petals were closed, 
5) Fewer than all 4 flat-field lamps were on,
6) More than 10000 pixels were rejected during fiber tracing, 
7) Fiber tracing fails, or traces are seperated by less than 3 pixels anywhere
on the flat-field image.

The first very critical step is to robustly determine the fiber positions 
in CCD pixel coordinates.
This is performed by identifying peaks in a cross-sectional cut of across 
the center row (actually a median cross-section of the central 9 rows).  
The correct identification of the 320 fibers per frame can be 
uniquely determined by demanding a wider spacing between the bundles 
of 20 fibers (spaced by about 10 pixels) relative to the nominal inter-fiber 
spacing (6.1 pixels on average).  Once the starting peaks and 
identification have been determined, we trace the fiber centers 
by performing flux-weighted centroids on subsequent rows
using the previous row's center as the intial guess.  This has proven to 
work well on SDSS flat-fields, and we have yet to uncover a failure 
mode in over 10000 flat-field frames.
The flux-weighted centroids do have scatter from shot noise, 
so we fit a 7th order polynomial to each set of fiber centroids, 
and the best-fit coefficients are stored to represent the fiber positions 
on each flat-field frame.

Each flat-field frame is extracted twice, once assuming a Gaussian
profile and again assuming the steeper $|x|^3$ profile.  In both cases,
the width is allowed to vary freely.  The total number of free parameters
per row in the flat-field extractions is equal to 2 per fiber plus 10
parameters to represent a smooth polynomial background.  After the extractions
are complete, $\chi^2$ is computed by directly comparing the model fit
to the flat-field image.  The median $\chi^2$ for each case is compared
directly, and the better fit and profile (lower median $\chi^2$) are retained
and implemented for the extraction of the remaining frames.  Also retained
is a smooth polynomial fit the average profile width in each fiber bundle
as a function of CCD row number.

After all flat-field frames for a given sequence have been extracted
and stored, the arc-lamp frames in the same sequence are extracted.  
Each arc-lamp frame is matched with the nearest (in time) 
non-rejected flat-field frame.  The matched flat-field gives the 
initial trace centers to
be assumed, as well as the profile shape and width determined from the
flat-field.  To account for any flexure between the frames, the trace
centers are allowed to shift/stretch to the arc-lamp image given a 
3x3 polynomial distortion kernel over the entire extent of the image
(match_trace.pro).

The arc-lamp frames are extracted row-by-row, in the same
manner as the flat-fields, the number of free parameters assumed
for the model fit is 2 per fiber (profile area and width) and 1 parameter
to fit for a constant background per row.  The restricted freedom in the
background parameterization reflects the lower scattered light levels in
the images of the arc-lamps.

\subsection{(or arc-lamp spectra -> wavelength map)}

We determine an initial wavelength solution for every arc-lamp frame
successfully extracted in the following manner:

For the appropriate side of each spectrograph (red or blue), we 
construct a list of suitable unblended lines from a parent list including
known lines of Mercury, Cadmium, Neon and (a trace of) Argon 
that are readily observable in the standard set of SDSS arc-lamps. 
A representative spectrum is calculated as the median extracted spectrum
of the 5 closest "unflagged" spectra falling near the central columns 
of the arc-lamp image.  This median arc-spectrum is matched with 
cross-correlation to a simulated spectrum of suitable arc-line positions and 
strengths by systematically searching a small volume of 
parameter space which conservatively covers
the polynomial coefficients of the SDSS wavelength solution.
In the first iteration, we search a set of 
100 coefficient vectors allowing for quadratic changes and return 
the best match.  In the second iteration, we search in a smaller region
about the previous best match, and search a list of 125 coefficient vector
allowing for cubic deviations.  The coefficient vector with the best
match from the second iteration is returned as an initial guess along with
the cross-correlation coefficient.

A 2-step iterative procedure is loop through
a range of possible polynomial
